Fail Faster

### 2025-03-15 paper/paper.tex
unpolished

Property-based testing (PBT) is a widely-used testing framework\hg{PBT isn't a
framework, it's a ``technique'' or ``approach'' --- Base\_QuickCheck is a framework} consisting of
two key components: a set of \textit{properties} that a system must satisfy, and
a way of producing
a large number of \textit{inputs} to that system. In contrast to traditional
unit testing, where test inputs are hand-written, PBT users create a random input
\textit{generator} that automatically produces inputs. Sometimes, this process
can itself be automated through the use of PBT libraries that synthesize
generators from type definitions; however, when inputs are required to maintain
invariants not communicated by the type, users must write a custom generator (or
filter out a broad class of invalid inputs).

\hg{I think this paragraph buries the lede. I'd start the paragraph talking
about performance and citing the 50ms number, and then jump right into what's
slow.}
Writing ``good'' generators—ones that successfully uncover bugs in the system under
test—is challenging. Much effort has been dedicated to this problem, often by
developing sophisticated domain-specific languages for specifying constrained
generators. However, these approaches do not always reflect the needs of real
PBT users. Under ideal conditions, a PBT suite would continue to run
until it stopped finding bugs, real-world usage patterns are very different; a
study of expert PBT users reported time budgets of between 50 milliseconds and
30 seconds for their test suites~\cite{inpractice}. Consequently, generation
time is a significant factor in PBT efficacy---that is, if a generator can
produce inputs twice as fast, it has twice as many chances within a given time
bound to find a bug. Therefore, it is imperative that PBT libraries are built
with performance in mind.

To better understand and address performance challenges in the landscape of PBT
generators, we explore popular libraries across several widely-used programming
languages.\hg{Again, this sentence also starts the paragraph too slowly; rather than
focus on our exploration (which isn't really the point of the paper) lead with
the result --- that abstraction overhead, boxing/unboxing, and allocations slow
down QuickCheck ports} We find that the flexibility of these libraries comes at a cost to 
performance,
introducing abstraction overhead, frequent boxing and unboxing, and avoidable
allocations (which lead to costly garbage collection pauses). We focus primarily 
on OCaml's \bq, selected for its efficiency relative to other OCaml PBT libraries 
and its integration with industrial build systems\hg{Let's not try to justify
our choice of OCaml \bq here --- we should just talk about strict functional
languages and justify this choice later}, where it is expected to run 
within tight time constraints. These
inefficiencies raise a fundamental question: how can PBT libraries produce code
that is both efficient \textit{and} general? By analyzing
\bq\ and similar libraries in other languages \tr{Rust?
More?}, we provide a broader perspective on common performance challenges in PBT
and explore potential strategies for more efficient generator implementations.

Our solution is \name, a generator library\hg{Can we say ``an approach to PBT
generator libraries'' or something more general like that?} that preserves
\bq’s functionality while improving performance through
\textit{multi-stage programming} (staging). Staging is a lightweight,
domain-specific compilation technique that allows us to completely eliminate the
runtime overhead of many common generator abstractions. \jwc{This relies on the fact that the code of generators is known statically at compile time} For instance, in
\name, monadic operations are zero-cost.
\name’s staged eliminates unnecessary allocations,
generates static control structures, and leverages unboxed integer libraries to
achieve performance improvements over \texttt{base\_quickcheck}, making it
well-suited to real-world PBT applications.

We \hg{also} present a new methodology for evaluating the relative effectiveness of PBT
generators.\hg{Might be an over-claim, we'll need to write this carefully}
Prior work has compared generators based on bug-finding ability: 
generators are considered equally effective if they find bugs in a
system at the same rate over a large number of trials ~\cite{etna}. However,
because \name\ preserves the semantics of
\bq, a \name\ generator produces exactly
the same inputs as its \bq\ counterpart. By comparing
generators based on \textit{program equality}, we ensure that any speedups in
bug-finding ability stem from performance enhancements rather than variations in
input distribution. This approach establishes a foundation for benchmarking
optimizations that enhance performance without compromising expressiveness.
\hg{I'm not totally convinced by this paragraph. Here's how I'd explain this:\\
In order to justify the switching cost associated with using \name, we designed
\name\ to be incredibly easy to benchmark: most PBT generator optimizations make
changes up to {\em distributional equivalence}, allowing the precise sampling
order to differ as long as the probability distribution is maintained, but
\name's optimizations are correct up to {\em program equivalence}. This means
that a \name\ generator produces precisely the same values in the same order as
the original generator, only faster. With this setup, benchmarking becomes
significantly easier than for prior work---our measurements have incredibly low
variance, and performance gains from \name\ translate directly to improved
bug-finding speed.}

Finally, we conduct an extensive evaluation, implementing both type-derived and
custom generators in \name\ and \bq. Our
generators produce a diverse range of inputs, including recursive data
structures, lambda calculus terms, and regular expressions. We benchmark
generation time, resource usage, and time-to-failure for each pair of staged and
unstaged generators, demonstrating that \name\ achieves
significant gains in speed and resource.

In summary, we dramatically improve the bug-finding effectiveness of PBT generators
by optimizing them for speed. Specifically, we make the following contributions:
\begin{enumerate}
    \item An empirical analysis of the sources of inefficiency in PBT
    generators. \jwc{And an argument that PBT generator libraries should be
    engineered for performance}
    \hg{Let's talk about this, I'm not sure I buy it}
    \item A library, \name, which offers efficient \jwc{performant?} generator \proposechange{functions}{combinators} through the
    use of multi-stage programming; this is the first known application of
    staging to PBT.\hg{Let's also talk about this}
    \item The insight that generators should be compared by program equality
    \item An evaluation demonstrating the improved performance of generators constructed
    using our library in a controlled comparison.
\end{enumerate}

@@

\section{Evaluation}
\subsection{Implementing and Testing Generators}
\label{subsection:pbt-for-pbt}
\jwc{Talk about the difftesting to ensure staging maintains semantics, and also different RNGs have identical semantics.}

>>>>

Property-based testing (PBT) uses generators of random data to produce inputs to a
system under test, which is then checked against a set of properties the system is
 expected to satisfy.  In practice, PBT is used as a lightweight, ad hoc approach 
 to software correctness, aimed at detecting bugs within a short testing 
 window---typically 50 milliseconds to 30 seconds. Given this practical constraint, 
 it is imperative that inputs be generated as fast as possible; in other words, 
 generators must be treated as performance-sensitive code.

 The locus for improvement in this area is \textit{generator libraries}: if programmers are 
 provided with the tools to write fast generators, then they do not themselves 
 need to be performance experts. However, existing libraries fail to achieve good performance 
 due to two major sources of inefficiency. The first is \textit{random sampling}, or the process
 of selecting random values. (Another sentence here?)
 The second is abstraction: generator libraries are implemented as embedded DSLS, which are flexible
 and expressive, but also introduce significant overhead that is not necessarily optimized away by 
 the compiler. These costs compound in the aggregate, raising a fundamental
 question: can generator libraries produce code that is both expressive \textit{and} efficient?

  % in strict functionallanguages like OCaml and Scala
We outline a fully general, two-stage approach to designing generator libraries, 
which provide high-level combinator functions with dramatically reduced 
overhead. First, \rand choice matters. We demonstrate that using a faster \rand
significantly impacts generator performance by directly optimizing an existing RNG, which
produces the same values as its unoptimized counterpart---but faster--thereby allowing us
to compare the RNG's impact on an existing generator library. Second, we 
describe a method of optimizing generator libraries using \textit{staged metaprogramming}, 
or ``staging''. Staged metaprogramming allows for code generation, which can mitigate 
performance challenges introduced by embedded DSLs. In particular, when a generator DSL 
introduces overhead for dynamic control flow, staging can specialize that control flow
and compile it statically. For example, when choosing between a list of options, a staged
generator can avoid allocating the list of options at runtime by expanding to an 
if-else. This approach is not tied to any particular PBT library---instead, it gives a recipe
for improving even SOTA PBT libraries.

To demonstrate
the effectiveness of our technique, we apply our technique to a leading PBT library, \bq, built with performance engineering in mind.
Our version of this library, termed \name, is a direct drop-in replacement for the original that is completely
semantically equivalent: generators written using it produce exactly the same values as 
those written in the original version. However, in \name, the abstractions provided by the generator DSL are 
zero-cost, and the \rand is highly-optimized. The direct equivalence between \name and \bq
allows us to perform extremely fine-grained comparisons between generators---generators implemented with both versions of the
library will produce exactly the same sequence of values, so differences in testing performance are directly attributable
to \name's performance speedup.

To test \name's effectiveness, we run a bunch of experiments to show that \name runs faster, allocates less, etc., than
\bq on a variety of different workloads. We demonstrate the generality of the technique by implementing \name in
Scala, and benchmarking the same generators in the Scala port of the library. Finally, as a sanity check, we demonstrate
that \name allows you to find bugs significantly faster ...

We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
    namely, \rand choice, and abstraction overhead. We describe a general procedure to address the latter concern based on staged metaprogramming. \secref{section:motiv}
    \item We implement a version of \bq, \name, that contains our optimizations and maintains strict program equality. \secref{section:4}
    \item We evaluate \name by demonstrating the improved performance of generators constructed
    using our library in a controlled comparison, as well as its impact on bug-finding ability. \secref{section:eval}
\end{enumerate}

@@

\section{Evaluation}
\label{section:eval}
\subsection{Implementing and Testing Generators}
\label{subsection:pbt-for-pbt}
\jwc{Talk about the difftesting to ensure staging maintains semantics, and also different RNGs have identical semantics.}
